---
title: "Appendix 2: Significance of Transfer Entropy coefficients"
toc: false

output:
  bookdown::pdf_book
  
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

Results for each permutation of age groups from each country. Tables include transfer entropy estimates, the effective transfer entropy estimates, standard errors and p-values based on bootstrap replications of the Markov chains under the null hypothesis of statistical independence.

```{r setup, include = FALSE}
# libraries --------------------------------------------------------------------
library(here)
library(tidyverse)
library(knitr)
library(kableExtra)

# options ----------------------------------------------------------------------
set.seed(123) # Seed for random number generation
options(scipen = 999) # disable sci number format
opts_chunk$set(
  cache.extra = rand_seed, 
  cache = FALSE,
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  echo = FALSE, 
  fig.retina = 3,
  dev = "ragg_png"
  )
```

```{r}
# variables --------------------------------------------------------------------
age_order <- c(
  "aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24",
  "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+"
)

# data -------------------------------------------------------------------------
te_results <- read_csv(here("data/te_results.csv")) |> 
    mutate(
      X = factor(X, levels = age_order),
      Y = factor(Y, levels = age_order)
    ) |> 
  select(-direction)
```

```{r}
te_results |>
  filter(X == "aged0to4") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged0to4") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged5to9") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged5to9") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage
```{r}
te_results |>
  filter(X == "aged10to14") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged10to14") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged15to19") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged15to19") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged20to24") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged20to24") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged25to34") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged25to34") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged35to44") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged35to44") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged45to54") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged45to54") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged55to64") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged55to64") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged65+") |> 
  filter(Y %in% c("aged0to4", "aged5to9", "aged10to14", "aged15to19", "aged20to24")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```

\newpage

```{r}
te_results |>
  filter(X == "aged65+") |> 
  filter(Y %in% c("aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65+")) |> 
  arrange(Y) |> 
  kbl(
    caption = "Significance of Transfer Entropy coefficients",
    digits = 2
  ) |> 
  kable_classic(full_width = F) |> 
  collapse_rows(columns = 1:2, valign = "middle") |> 
  kable_styling(font_size = 7)
```
